Windowed residual-minimization model reduction for dynamical systems

### 2019-10-02 main.tex
latest

\KTC{Why are we reviewing LSFEM? This isn't a ROM. Don't think it's relevant}
Least-squares residual minimization as a numerical solution technique for
partial differential equations was pioneered in the finite-element community
and led to what is known as the least-squares finite-element method
(LSFEM)~\cite{ls_review,bochev_leastsquares}. 
\KTC{Specifically, LSFEM was designed to solve the inf--sup stability problem.
Be more precise.}
Motivated by the fact that the
Galerkin method can be inaccurate and unstable for nonlinear and
non-self-adjoint systems, the LSFEM involves (1.) forming a quadratic
functional comprising the residual norm of the partial differential equation
of interest and (2.) leveraging variational principles to derive the
stationary conditions of the quadratic functional; i.e, LSFEM can be thought
of as finding the solution within a finite element trial space the minimizes
the $2$-norm of the PDE residual. Least-squares methods often exhibit numerous
advantages (e.g., least-squares system are symmetric and positive definite)
\KTC{This is not an `advantage'. It is guaranteed to be inf--sup stable.}
over the more standard Galerkin discretization.

Researchers also developed residual-minimization methods for model reduction.
Similar to LSFEM, residual minimization in this context associates with
computing the solution within a low-dimensional trial space that minimizes the
$\ell^2$-norm of the full-order-model residual. \KTC{haven't defined
`full-order model'} Researchers have developed such residual-minimizing
model-reduction methods for both static systems (i.e., systems without
time-dependence)
\cite{legresley_1,legresley_2,legresley_3,bui_resmin_steady,rovas_thesis,carlberg_thesis,bui_thesis}
and dynamical
systems~\cite{bui_thesis,bui_unsteady,carlberg_thesis,carlberg_gnat,carlberg_lspg,carlberg_lspg_v_galerkin}.
In the latter category,
Refs.~\cite{bui_thesis,bui_unsteady,carlberg_thesis,carlberg_gnat,carlberg_lspg}
formulated the residual minimization problem for dynamical systems by
sequentially minimizing the \textit{time-discrete} full-order-model residual
(i.e., the residual arising after applying time discretization) at each time
instance on the time-discretization grid. This formulation is often referred
to as the \textit{least--squares Petrov--Galerkin} (LSPG) method.
Ref.~\cite{carlberg_lspg_v_galerkin} performed detailed analyses of this
formulation and examined its connections with Galerkin projection. Critically,
this work demonstrated that (1) under certain conditions, LSPG can be derived
from a Petrov--Galerkin projection executed on the time-continuous
full-order-model residual, and (2) LSPG and Galekrin projection are equivalent
in the limit as the time step goes to zero (i.e., Galerkin projection
minimizes the time-instantaneous full-order model residual). Numerous
numerical experiments have demonstrated that LSPG projection often yields more
accurate and stable solutions than Galerkin
projection~\cite{bui_thesis,carlberg_lspg_v_galerkin,carlberg_gnat,carlberg_thesis,parish_apg}.
The common intuitive explanation for this improved performance is that, by
minimizing the full-order-model residual over a finite time window (rather
than time instantanously), LSPG projection computes solutions that are more
accurate over a larger part the trajectory compared with Galerkin projection.

@@

The LSPG approach sequentially performs residual minimization over a single
time-step, and as such is not associated with residual minimization over the
entire time interval. \KTC{This is what ST-LSPG does. Why do we ignore it? the
full dynamical system} Indeed, LSPG is not equipped with \textit{a priori}
guarantees of stability or convergence, even for linear time-invariant
systems~\cite{bui_thesis}. Further, the performance of LSPG depends on the
time step in two distinct ways: changing the time-step size modifies both the
time window over which LSPG minimizes the residual \textit{as well as} the
time-discretization error of the full-order model on which LSPG is based. As a
result, LSPG often yields the smallest error for an intermediate value of the
time step (see, e.g., Ref.~\cite[Figure 9]{carlberg_lspg_v_galerkin}). In
particular, as the time step decreases, the accuracy of LSPG approaches that
of Galerkin (which is often poor)\footnote{This occurs because LSPG projection
is equivalent to Galerkin projection in the limit $\Delta t \rightarrow 0$.};
as the time step increases, the time discretization error of the full-order
model (on which the LSPG ROM is based) increases.  LSPG projection additionally is
additionally sensitive to the time discretization scheme. When used with
explicit temporal discretization, for instance, LSPG again reverts to
Galerkin. To gain the enhanced stability associated with LSPG, the underlying
time-discretization scheme must be implicit with an appropriate time-step.
This can increase the cost of the reduced-order model, especially in the case
where the optimal time-step required by LSPG is small.

@@

This work seeks to overcome the practical challenges associated with existing
residual minimization approaches for reduced-order models. To this end, we
propose the \methodNameLower\ (\methodAcronym) method for model reduction,
which minimizes the FOM ODE residual over a sequence of arbitrarily defined
time windows. The key novelties of the work are twofold.  First, the method
minimizes the \textit{time-continuous}  residual (i.e., that associated with
the full-order model ODE). By adopting a time-continuous viewpoint, the
formulation decouples the underlying temporal discretization scheme from the
residual minimization problem, thus addressing a key deficiency of LSPG
\KTC{This is also true for ST-LSPG and maybe other discretization-based
methods? Have to check}. Time-continuous residual minimization also exposes
two different solution methods,  namely a \textit{discretize-then-optimize}
(i.e., direct) method and an \textit{optimize-then-discretize} (i.e.,
indirect) method. The second novelty is that the method sequentially minimizes
the residual over arbitrarily defined \textit{time windows} rather than
sequentially minimizing the residual over time steps or over the entire time
domain. This equips the method with additionaly flexibility that enables it to
explore more fine-grained tradeoffs between computational cost and error. We
demonstrate that this method can be viewed as a generalization of existing
methods, because particular instances of this formulation are equivalent to
Galerkin, LSPG, and space--time LSPG projection. Figure~\ref{fig:flowchart}
depicts how the proposed \methodAcronym\ method is related to existing
approaches.

>>>>

%\KTC{Why are we reviewing LSFEM? This isn't a ROM. Don't think it's relevant}
%Least-squares residual minimization as a numerical solution technique for
%partial differential equations was pioneered in the finite-element community
%and led to what is known as the least-squares finite-element method
%(LSFEM)~\cite{ls_review,bochev_leastsquares}. 
%\KTC{Specifically, LSFEM was designed to solve the inf--sup stability problem.
%Be more precise.}
%Motivated by the fact that the
%Galerkin method can be inaccurate and unstable for nonlinear and
%non-self-adjoint systems, the LSFEM involves (1.) forming a quadratic
%functional comprising the residual norm of the partial differential equation
%%of interest and (2.) leveraging variational principles to derive the
%stationary conditions of the quadratic functional; i.e, LSFEM can be thought
%of as finding the solution within a finite element trial space the minimizes
%the $2$-norm of the PDE residual. Least-squares methods often exhibit numerous
%advantages (e.g., least-squares system are symmetric and positive definite)
%\KTC{This is not an `advantage'. It is guaranteed to be inf--sup stable.}
%over the more standard Galerkin discretization.

Residual minimization methods in model-order reduction comprise 
computing a solution within a low-dimensional trial space that minimizes some residual norm of the system of interest. Researchers have developed such residual-minimizing
model-reduction methods for both static systems (i.e., systems without
time-dependence)
\cite{legresley_1,legresley_2,legresley_3,bui_resmin_steady,rovas_thesis,carlberg_thesis,bui_thesis}
and dynamical
systems~\cite{bui_thesis,bui_unsteady,carlberg_thesis,carlberg_gnat,carlberg_lspg,carlberg_lspg_v_galerkin}.
In the latter category,
Refs.~\cite{bui_thesis,bui_unsteady,carlberg_thesis,carlberg_gnat,carlberg_lspg}
formulated the residual minimization problem for dynamical systems by
sequentially minimizing the \textit{time-discrete} full-order-model residual
(i.e., the residual arising after applying time discretization) at each time
instance on the time-discretization grid. This formulation is often referred
to as the \textit{least--squares Petrov--Galerkin} (LSPG) method.
Ref.~\cite{carlberg_lspg_v_galerkin} performed detailed analyses of this
formulation and examined its connections with Galerkin projection. Critically,
this work demonstrated that (1) under certain conditions, LSPG can be derived
from a Petrov--Galerkin projection executed on the time-continuous
full-order-model residual, and (2) LSPG and Galekrin projection are equivalent
in the limit as the time step goes to zero (i.e., Galerkin projection
minimizes the time-instantaneous full-order model residual). Numerous
numerical experiments have demonstrated that LSPG projection often yields more
accurate and stable solutions than Galerkin
projection~\cite{bui_thesis,carlberg_lspg_v_galerkin,carlberg_gnat,carlberg_thesis,parish_apg}.
The common intuitive explanation for this improved performance is that, by
minimizing the full-order-model residual over a finite time window (rather
than time instantanously), LSPG projection computes solutions that are more
accurate over a larger part of the trajectory as compared to Galerkin projection.

@@

The LSPG approach has several shortcomings. First, the performance of LSPG depends on the
time discretization. Changing the time-step size modifies both the
time window over which LSPG minimizes the residual \textit{as well as} the
time-discretization error of the full-order model on which LSPG is based. As a
result, LSPG often yields the smallest error for an intermediate value of the
time step (see, e.g., Ref.~\cite[Figure 9]{carlberg_lspg_v_galerkin}). In
particular, as the time step decreases, the accuracy of LSPG approaches that
of Galerkin (which is often poor)\footnote{This occurs because LSPG projection
is equivalent to Galerkin projection in the limit $\Delta t \rightarrow 0$.}. 
%The accuracy of LSPG additionally degrades if the time-step size becomes too large. It is unclear if this is due to the discretization error associated with enlarging the time-step size, or rather if it is due to the size of the window the residual is being minimized over. LSPG projection additionally is
%additionally sensitive to the time discretization scheme. When used with
%explicit temporal discretization, for instance, LSPG again reverts to
%Galerkin. To gain the enhanced stability associated with LSPG, the underlying
%time-discretization scheme must be implicit with an appropriate time-step.
%%This can increase the cost of the reduced-order model, especially in the case
%where the optimal time-step required by LSPG is small. 
Second, as the LSPG approach sequentially performs residual minimization over
time-steps, is not equipped with any notion of optimality over the entire 
time domain of interest. As a result, LSPG is not equipped with \textit{a priori}
guarantees of stability or convergence, even for linear time-invariant
systems~\cite{bui_thesis}.

Researchers have pursued the development of space--time residual minimization ROMs~\cite{choi_stlspg,constantine_strom,URBAN2012203,Yano2014ASC,bui_thesis} to address this last issue. In existing space--time methods, the spatial \textit{and} temporal dependence of the state is represented in a low-dimensional space--time trial space. Space--time residual minimization approaches then comprise computing a space--time solution within this low-dimensional space--time trial space that minimizes the space--time norm of the full-order model. Space--time methods are thus equipped with a notion of optimality over the entire time domain of interest and, as a result, contain more favorable error bounds than the Galerkin and LSPG approaches.  Several drawbacks of existing space--time ROMs are that they (1) require minimizing the entire space--time residual (which can be a computationally intensive process) and (2) require the specification of a finite-dimensional temporal basis (computed, e.g., via POD applied to space--time snapshots). This latter issue is extremely relevant to the present work.  
%These space--time In these space--time ROMs, the temporal dependence of the state 
%is represented via a finite dimensional basis representation. 
%Space--time residual minimization methods can be prohibitively expensive, however, as they involve minimizing the entire space--time residual. Further, existing space--time approaches require a finite-dimensional space--time basis.

This work seeks to overcome the practical challenges associated with existing
residual minimization approaches for reduced-order models. To this end, we
propose the \methodName\ method for model reduction, which minimizes the FOM ODE residual
over a sequence of time windows. The key novelties of the work are two fold.
First, the residual minimization problem is formulated to minimize the
\textit{time-continuous} residual. By working from a time-continuous viewpoint, the
present formulation decouples the underlying temporal discretization scheme
from the residual minimization problem; thus addressing a key deficiency of
LSPG \KTC{This is also true for ST-LSPG and maybe other discretization-based
methods? Have to check}. The time-continuous viewpoint additionally exposes
two different methods that
can be used to solve the residual-minimization problems, namely
a \textit{discretize-then-optimize} (i.e., direct) method and
an \textit{optimize-then-discretize} (i.e., indirect) method. The second novelty
is that the residual is minimized over \textit{time windows} rather than over
a single time step or the entire time domain. This equips the method with
additionaly flexibility that enables it
to explore more fine-grained tradeoffs in computational cost and error. We
demonstrate that this method can be viewed as a generalization of existing
methods, as particular instances of this formulation recover Galerkin, LSPG,
and space--time LSPG projection. Figure~\ref{fig:flowchart} depicts how the
proposed method is related to existing approaches.

% \KTC{This is what ST-LSPG does. Why do we ignore it? the
%full dynamical system} Indeed, LSPG is not equipped with \textit{a priori}
%guarantees of stability or convergence, even for linear time-invariant
%systems~\cite{bui_thesis}. Further, the performance of LSPG depends on the
%time step in two distinct ways: changing the time-step size modifies both the
%time window over which LSPG minimizes the residual \textit{as well as} the
%time-discretization error of the full-order model on which LSPG is based. As a
%result, LSPG often yields the smallest error for an intermediate value of the
%time step (see, e.g., Ref.~\cite[Figure 9]{carlberg_lspg_v_galerkin}). In
%particular, as the time step decreases, the accuracy of LSPG approaches that
%of Galerkin (which is often poor)\footnote{This occurs because LSPG projection
%is equivalent to Galerkin projection in the limit $\Delta t \rightarrow 0$.};
%as the time step increases, the time discretization error of the full-order
%model (on which the LSPG ROM is based) increases.  LSPG projection additionally is
%additionally sensitive to the time discretization scheme. When used with
%explicit temporal discretization, for instance, LSPG again reverts to
%Galerkin. To gain the enhanced stability associated with LSPG, the underlying
%time-discretization scheme must be implicit with an appropriate time-step.
%This can increase the cost of the reduced-order model, especially in the case
%where the optimal time-step required by LSPG is small.

@@
